---
title: "NRSG 741 - Homework 2 - Exploratory Data Analysis"
author: "Melinda Higgins"
date: "February 25, 2020"
output: html_document
---

## INSTRUCTIONS

* Use this Rmarkdown file `N741Spring2020_Homework02.Rmd` to get started.
* Change the author to YOUR NAME
* Change the date
* Note: This Rmarkdown file has one R code chunk at the top that reads in the dataset and loads the R packages you will need.
* After each question below, insert an R code chunk to enter the R code needed to answer that question. Do this for each question.
* Outside of the R code chunk, type in any text needed to provide explanation or answer the questions further.

```{r setup, include=FALSE}
# set up chunk output options
# set to FALSE to clean up final output
# leave TRUE to help with debugging initially
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = TRUE)
knitr::opts_chunk$set(message = TRUE)
knitr::opts_chunk$set(error = TRUE)

# Get data
# learn more at https://archive.ics.uci.edu/ml/datasets/abalone

# Load the abalone dataset using read_csv() function
# from readr package
library(readr)
abalone <- readr::read_csv(file = "https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data",
                           col_names = FALSE)

# put in the variables names
# see details at https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names
names(abalone) <- c("sex","length",
                    "diameter","height",
                    "wholeWeight","shuckedWeight",
                    "visceraWeight","shellWeight","rings")

# Load dplyr and ggplot2 packages
library(dplyr)
library(ggplot2)
```

## Goal of Homework 2

This homework is meant to further your `dplyr` and `ggplot2` skills. 

## Modify R code chunks

In each of the R code chunks below, scaffolding is provided. Everywhere you see 3 underscores ___ , you will need to fill in the appropriate code, variable name, function name, etc.

## Abalones Dataset from UCI Repository

For this homework, you will keep working with the `abalone` dataset from the UCI data repository at [https://archive.ics.uci.edu/ml/datasets/abalone](https://archive.ics.uci.edu/ml/datasets/abalone).

Use tools within the `dplyr` package as much as possible to answer the following questions.

##### **Question 1:** What kind of R object is the `abalone` dataset?

```{r q1, eval=TRUE}
# insert R code here to answer question 1
# HINT: The name of the dataset is abalone

class(abalone)
```

TYPE ANSWER TO QUESTION 1 HERE

The `abalone` dataset is a `data.frame` - it is also a tibble data frame (`spec_tbl_df` and `tbl_df` and `tbl`).

##### **Question 2:** How many observations are in the `abalone` dataset?

```{r q2, eval=TRUE}
# HINT: there are multiple ways to answer this - pick one
# dim(___)
# str(___)
# glimpse(___)
# nrow(___)

dim(abalone)
str(abalone)
glimpse(abalone)
nrow(abalone)
```

TYPE ANSWER TO QUESTION 2 HERE

The number of rows (which is the number of observations) in the `abalone` dataset is `r nrow(abalone)`.

##### **Question 3:** For diameter, how many abalones have diameters less than 0.5mm?

```{r q3, eval=TRUE}
# the variable name is diameter
# abalone %>%
#   filter(___ < ___) %>%
#   nrow()

abalone %>%
  filter(diameter < 0.5) %>%
  nrow()
```

TYPE ANSWER TO QUESTION 3 HERE

The number of abalones in the dataset with diameter < 0.5 mm is 3388.

##### **Question 4:** How many abalones have shucked weights larger than their whole weight? 

NOTE: There should be NO measurements where the shucked weight is > whole weight. If there are some these are probably data entry errors in this dataset.

```{r q4, eval=TRUE}
# HINT: Use a logical expression inside a filter step
# HINT: Check the spelling and case for the 
# variable names for shucked weight and whole weight
# abalone %>%
#   filter(___ > ___) %>%
#   nrow()

abalone %>%
  filter(shuckedWeight > wholeWeight) %>%
  nrow()
```

TYPE ANSWER TO QUESTION 4 HERE

The number of abalones which have shucked weights recorded as larger than > their whole weights is 4.

---

Create a subset containing only infants `sex == "I"`. Call this new dataset `infants`

```{r infants, eval=TRUE}
# HINT: Put the logical statement inside the filter() function
# Dont forget to use the assign operator <- to create the infants object
# infants <- abalone %>%
#   filter(___ == ___)

infants <- abalone %>%
  filter(sex == "I")
```

##### **Question 5:** How many infants are in this subset?

```{r q5, eval=TRUE}
# Hint: see code in question 2 above
# pick the function you prefer to answer this question
# ___(___)

nrow(infants)
```

TYPE ANSWER TO QUESTION 5 HERE

The number of infant abalones in this dataset are `r nrow(infants)`.

---

Show off your `dplyr` skills with `group_by()` - we didn't get a chance to fully explore `group_by()` in class but it is added in the examples below to help you answer these questions.

##### **Question 6:** What is the average whole weight for each abalone sex (get whole weight means for females "F", males "M" and infants "I" separately)?

```{r q6, eval=TRUE}
# Hint: put the variables used in the select statement
# and in the summarise() statement. Remember to put
# in a name for the output of the mean() function
# something like meanwt
# abalone %>%
#   select(sex, ___) %>%
#   group_by(sex) %>%
#   summarise(___ = mean(___, na.rm=TRUE))

abalone %>%
  select(sex, wholeWeight) %>%
  group_by(sex) %>%
  summarise(meanwt = mean(wholeWeight, na.rm=TRUE))

```

TYPE ANSWER TO QUESTION 6 HERE

The output above shows the average whole weight of the abalones by sex (males, females and infants).

##### **Question 7:** Get the means for the abalone length and height by sex?

```{r q7, eval=TRUE}
# Hint: put variable names in the select statement
# put the function name for the mean in the 
# summarise_all() function
# abalone %>%
#   select(sex, ___, ___) %>%
#   group_by(sex) %>%
#   summarise_all(___, na.rm=TRUE)

abalone %>%
  select(sex, length, height) %>%
  group_by(sex) %>%
  summarise_all(mean, na.rm=TRUE)
```

TYPE ANSWER TO QUESTION 7 HERE

The table above provides the average length and height for the abalones by sex.

----

## Test your graphing skills using `ggplot2`

Using the `abalone` dataset, create the following graphics/figures using `ggplot()` and associated `geom_xxx()` functions.

##### **Question 8:** Create a histogram of abalone whole weight

BONUS: Outline the histogram bars with a black line and fill the histogram bars with a green color

```{r q8, eval=TRUE}
# Hint: the first option in the ggplot() function 
# is the name of the dataset
# the variable name is put inside aes()
# ggplot(___, aes(___)) +
#   geom_histogram()

# BONUS:
# ggplot(___, aes(___)) +
#   geom_histogram(color = ___, fill = ___)

ggplot(abalone, aes(wholeWeight)) +
  geom_histogram()

# BONUS:
ggplot(abalone, aes(wholeWeight)) +
  geom_histogram(color = "black", fill = "green")
```

_What do you notice about the distribution (any outliers or skewness)?_

TYPE ANSWER TO QUESTION 8 HERE

The distribution is slightly right skewed (longer tail to the right) and the curve is slightly fatter than would be expected under a normal curve indicating some slight negative kurtosis (flatter or fatter than a normal curve).

##### **Question 9:** Create side-by-side boxplots of the number of rings by gender - color the bars by sex

_HINT use `geom_boxplot` with x = sex and y = rings_ 

```{r q9, eval=TRUE}
# ggplot(___, 
#        aes(x=___, y=___, fill=___)) + 
#   geom_boxplot()

ggplot(abalone, 
       aes(x=sex, y=rings, fill=sex)) + 
  geom_boxplot()
```


##### **Question 10:** Create a scatterplot of the whole weight on the X axis and shucked weight on the Y axis and color the points by sex

```{r q10, eval=TRUE}
# ggplot(___, 
       # aes(x=___, y=___, color=___)) + 
  # geom_point() + 
  # geom_abline(intercept = 0, slope = 1)

ggplot(abalone, 
       aes(x=wholeWeight, y=shuckedWeight, color=sex)) + 
  geom_point() + 
  geom_abline(intercept = 0, slope = 1)
```

Can you see which abalones have shucked weights > whole weights which should not happen? Look at the y=x reference line. What sex are the abalones with the incorrect weights?

From the plot above, it appears that there are 4 abalones that have shucked weights > their whole weight and all 4 are infants (green points).

## Final Instructions

* KNIT this RMD file to PDF (or to HTML or DOC and save as PDF)
* Upload your PDF document to Canvas
